Temperature Effects on Surface Sulfur Dioxide

### 2011-07-26 Background.tex
Minor edits before sending to Baer

The optimized geometry of the \suldiox-hydrates suggest cyclic bonding structures, but it remains a different entirely when a \suldiox~is placed in a dynamics environment such as in the course of molecular dynamics simulations of an aqueous surface. Although geometry optimization shows the formation of these cyclic complexes, do they form during a dynamic bonding process on a simulated water surface? Graph theoretical techniques can be employed to determine if cyclic structures form during the simulations. Previous use of graphs in molecular computations were applied to finding stable arrangements of water clusters, ice, hydrogen bonding, extracting topological molecular properties, and cyclic structure studies.\cite{Anick2002, Huber2007, Radhakrishnan1991, Shi2005, Garcia2004, McDonald1998}

@@

Baer et al. presented a detailed geometric and spectroscopic breakdown of types \Rmnum{3} cycles with two and three waters from their DFT calculations.\cite{Baer2010} Given the information of the number of waters, atoms, and bonds involved in the bonding cycles, we find that of the three-water type \Rmnum{3} cycles, there exist two structural varieties, shown in Figure \ref{fig:type-3-varieties}, that differ in the arrangement of the waters involved. Type \Rmnum{3}A is arranged with each water contributing an OH bond to the structure of the cycle. Type \Rmnum{3}B involves a single water contributing an OH, whereas the other two waters contribute only the oxygen atom or the entire water molecule to the structure, respectively. This nuance of the three-water type \Rmnum{3} structures, and the overall distribution are presented in more detail later. We also show the distribution of cyclic structures encountered during MD simulations to further understand the behaviors of \suldiox-hydrates at the water surface.

>>>>

The optimized geometry of the \suldiox-hydrates suggest cyclic bonding structures, but it remains a different story entirely when a \suldiox~is placed in a dynamic environment such as in the course of molecular dynamics simulations of an aqueous surface. Although geometry optimization shows the formation of these cyclic complexes, do they form during a dynamic bonding process on a simulated water surface? Graph theoretical techniques can be employed to determine if cyclic structures form during the simulations. Previous use of graphs in molecular computations were applied to finding stable arrangements of water clusters, ice, hydrogen bonding, extracting topological molecular properties, and cyclic structure studies.\cite{Anick2002, Huber2007, Radhakrishnan1991, Shi2005, Garcia2004, McDonald1998}

@@

Baer et al. presented a detailed geometric and spectroscopic breakdown of type \Rmnum{3} cycles with two and three waters from their DFT calculations.\cite{Baer2010} Given the information of the number of waters, atoms, and bonds involved in the bonding cycles, we find that of the three-water type \Rmnum{3} cycles, there exist two structural varieties, shown in Figure \ref{fig:type-3-varieties}, that differ in the arrangement of the waters involved. Type \Rmnum{3}-A (shown in Figure \ref{fig:type-3-varieties}A) is arranged with each water contributing an OH bond to the structure of the cycle. Type \Rmnum{3}B involves a single water contributing an OH, whereas the other two waters contribute only the oxygen atom or the entire water molecule to the structure, respectively. This nuance of the three-water type \Rmnum{3} structures, and the overall distribution are presented in more detail later. We also show the distribution of cyclic structures encountered during MD simulations to further understand the behaviors of \suldiox-hydrates at the water surface.

### 2011-07-26 Coordination-Results.tex
Minor edits before sending to Baer

The bonding coordination of the surface \suldiox~was determined at each timestep of the simulations. Figure \ref{fig:bonding-coordinations} shows the distribution of coordinations of the surface \suldiox, by percent of simulation time, across all the simulations for both the cold (blue) and hot (red) trajectories. A first visual inspection clearly indicates the two most common coordinations: ``S'' and ``SO'', followed by the unbound coordination, and ``SOO''. This trend holds at both temperatures. Also, a \suldiox~configuration with bonds forming solely through the oxygens without any sulfur interactions, or any configuration with more than a single interaction the the sulfur atom are rare in comparison to the single sulfur coordinations. These less common coordinations likely form as transitory configurations when the waters briefly move closer or further away from the surface \suldiox.

The distribution paints a likely picture of the surface adsorption process of a \suldiox~first coming in contact with surface waters. Starting in an unbound state, two methods of initial contact are possible: through the \suldiox-sulfur of through one of the oxygens. Because the population of non-sulfur interactions (i.e. ``O'', ``OO'', ``OOO'', etc.) only account for 3\% of all the configurations sampled, either there is a very brief transition state where the \suldiox~first binds through an oxygen and then quickly adds, or switches to a sulfur interaction with the surface waters, or the first contact is made through the sulfur atom. Because the ``S'' coordination has such a large population, accounting for 35\% of the trajectories in the cold systems, and 32\% in the hot systems, it is most likely that the first binding contact to the waters is through the \suldiox~sulfur to a water oxygen. Furthermore, once in the ``S'' coordination, the \suldiox~can either unbind, add a second sulfur interaction, or bind through an oxygen. The histogram of Figure \ref{fig:bonding-coordinations} suggests that the preferred route is to further coordinate through an oxygen to achieve the ``SO'' coordination. ``SO'' accounts for 35\% of the cold coordinations, and 36\% of the hot trajectories. Binding to a second oxygen to the ``SOO'' coordination (7\% of the cold, 12\% of the hot systems) is less likely than losing an interaction through an oxygen back to the ``S'' coordination.

@@

Baer et al performed this coordination analysis for their single-temperature study, but discriminating between \suldiox~binding through the two different oxygens. They concluded that there is asymmetric hydrogen bonding through the \suldiox~oxygens, with one binding more often than the other. This is supported by the findings here where all the double oxygen coordinations (i.e. ``OO'', ``SOO'', etc.) represent a much lower percentage of the coordinations than the single oxygen counterparts (i.e. ``O'', ``SO'', etc.).

@@

The binding behavior of the \suldiox~is altered by changing the temperature of the system, as evidenced in the shift in coordination populations of Figure \ref{fig:bonding-coordinations} from cold to hot. In the cold temperature, the unbound, single oxygen, and ``S'' coordinations are more populated than in the hot systems. The increased temperature causes all the coordinations with a single sulfur interaction or more (with the exception of ``S'') to increase over the equivalent cold temperature coordination populations. We see that the cold \suldiox~spends nearly twice as much time unbound, and only slightly more time in the ``S'' coordination. Most of the unbound population in the hot simulations appears to have shifted to the states with double-oxygen and double-sulfur coordinations. Figure \ref{fig:coordination-breakdown} shows the distribution of the number of bonds through the \suldiox~sulfur (top) and through the \suldiox~oxygens (bottom). When comparing from cold to hot distributions, the greatest shift occurs in the number of \suldiox~oxygen bonds as the hot \suldiox~has more bonding to the oxygens. The increased temperature allows the \suldiox~oxygens to bind to more water hydrogens than in the cold systems.

@@

These increased binding through the \suldiox~oxygens at the hot temperatures should have a visible effect on the radial distribution function (RDF) between \suldiox-oxygen and water-hydrogen. Figure \ref{fig:rdf} shows the RDF of the \suldiox~sulfur and oxygen atoms with their water bonding partners, water oxygen and hydrogen, respectively, at both cold (blue) and hot (green) temperatures. Both the first and second peaks of the O$_{SO_2}$-H$_{H_2O}$ RDF increase at the higher temperature, indicating a large number of water hydrogens near the \suldiox~oxygens in both hydration shells. This corroborates the increased coordination through the \suldiox~oxygens to the surface waters at the higher temperature.

@@

The populations of the single-oxygen coordinations are similar at both temperatures, somewhat contradicting the results of the RDF. We expect an increase of all \suldiox-oxygen interactions with surface water hydrogens, but the non-sulfur coordinations do not obey this trend. This indicates that the non-sulfur coordinations do not behave as most of the other coordinations, and their small contribution to the overall population in the distribution of Figure \ref{fig:bonding-coordinations} indicates that they are anomalous, or only briefly encountered during the simulations.

The same temperature effect of shifting population to higher coordinations also manifests in the \suldiox~sulfur. The single-``S'' coordinations have similar populations at both temperatures, but the change from cold to hot temperature shifts the unbound \suldiox~coordinations up towards to the doubly-bound sulfur coordinations (i.e. ``SS'', ``SSO'', etc.). The increased temperature causes more binding to the \suldiox~by allowing for multiple bonds through both the \suldiox~sulfur, and through the oxygens. Thus, the hotter temperature allows for increased coordination of the \suldiox~to the water surface.

When the \suldiox~is on the cold water surface, the ``S'' and ``SO'' coordination populations are nearly equally represented. As the temperature increases to the hot system, the ``S'' coordination decreases, while the ``SO'' increases. We see from the RDFs that colder temperatures favor binding through the \suldiox~sulfur because of the larger first peak in the S$_{SO_2}$-O$_{H_2O}$ RDF. From this increase we expect the ``S''-coordinated \suldiox~to be more prevalent at cold temperatures, as is the case for the coordination distribution. However, the O$_{SO_2}$-H$_{H_2O}$ RDF indicates increased hydrogen bonding through the \suldiox~oxygens at higher temperatures manifested by a larger first peak in Figure \ref{fig:rdf} for the hot temperature. Thus the minimally ``SO''-coordinated distributions (those with at least a single bond to an oxygen and the sulfur, i.e. ``SO'', ``SSO'', ``SSOO'', etc.) are more prevalent at the high temperature. Overall, we find evidence from both the bonding coordination distribution, and from the RDFs to suggest a modified bonding behavior at different temperatures. At an increased temperature the surface \suldiox~binds more through its oxygens than the cold counterpart. However, like the \suldiox~in the cold temperature, it likely has a persistent interaction through the sulfur atom to a neighboring water oxygen throughout most of the simulated trajectories.

>>>>

The bonding coordination of the surface \suldiox~was determined at each timestep of the simulations. Figure \ref{fig:bonding-coordinations} shows the distribution of bonding coordinations of the surface \suldiox, by percentage of all bonding coordinations encountered, across all the simulations for both the cold (blue) and hot (red) trajectories. A first visual inspection clearly indicates the two most common coordinations: ``S'' and ``SO'', followed by the unbound coordination, and ``SOO''. This trend holds at both temperatures. Also, a \suldiox~configuration with bonds forming solely through the oxygens without any sulfur interactions, or any configuration with more than a single interaction through the sulfur atom are rare in comparison to the single sulfur coordinations. These less common coordinations likely form as transitory configurations when the waters briefly move closer or further away from the surface \suldiox.

The distribution paints a likely picture of the surface adsorption process of a \suldiox~first coming in contact with surface waters. Starting in an unbound state, two methods of initial contact are possible: through the \suldiox-sulfur or through one of the oxygens. Because the population of non-sulfur bonding interactions (not including the ``unbound'' state, i.e. ``O'', ``OO'', ``OOO'', etc.) only account for 3\% of all the configurations sampled, either there is a very brief transition state where the \suldiox~first binds through an oxygen and then quickly adds, or switches to a sulfur interaction with the surface waters, or the first contact is made through the sulfur atom. The ``S'' coordination has a large population, accounting for 35\% of the trajectories in the cold systems, and 32\% in the hot systems. Thus, it is most likely that the first binding contact to the waters is through the \suldiox~sulfur to a water oxygen. Furthermore, once in the ``S'' coordination, the \suldiox~can either unbind, add a second sulfur interaction, or bind through an oxygen. The histogram of Figure \ref{fig:bonding-coordinations} suggests that the preferred route is to further coordinate through an oxygen to achieve the ``SO'' coordination. ``SO'' accounts for 35\% of the cold coordinations, and 36\% of the hot trajectories. Binding to a second oxygen to the ``SOO'' coordination (7\% of the cold, 12\% of the hot systems) is less likely than losing an interaction through an oxygen back to the ``S'' coordination.

@@

Baer et al performed this coordination analysis for their single-temperature study, but discriminated between \suldiox~binding through the two different oxygens. They concluded that there is asymmetric hydrogen bonding through the \suldiox~oxygens, with one binding more often than the other. This is supported by the findings here where all the double oxygen coordinations (i.e. ``OO'', ``SOO'', etc.) represent a much lower percentage of the coordinations than the single oxygen counterparts (i.e. ``O'', ``SO'', etc.).

@@

The binding behavior of the \suldiox~is altered by changing the temperature of the system, as evidenced in the shift in bonding coordination populations of Figure \ref{fig:bonding-coordinations} from cold to hot. In the cold temperature, the unbound, single oxygen, and ``S'' coordinations are more populated than in the hot systems. We see that the cold \suldiox~spends nearly twice as much time unbound, and only slightly more time in the ``S'' coordination than in the hot system. The increased temperature of the hot systems causes all the coordinations with more than a single sulfur interaction (``SS'', ``SSO'', etc.) to increase over the equivalent cold temperature coordination populations. Most of the unbound population in the hot simulations appears to have shifted to the states with double-oxygen or double-sulfur coordinations. Figure \ref{fig:coordination-breakdown} shows the distribution of the number of bonds through the \suldiox~sulfur (top) and through the \suldiox~oxygens (bottom). When comparing from cold to hot distributions, the change in temperature increases the number of bonds to the \suldiox~sulfur and oxygen atoms. Whereas in the cold system the \suldiox~very rarely forms two interactions through the sulfur, in the hot system we see that the two-sulfur interaction bonding coordination account for nearly 5\% of all coordinations. There is a similar increase in bonding to the \suldiox~oxygens in the hot system. The increased temperature of the hot systems allows the \suldiox to bind to more surface waters than in the cold systems.

@@

The increased binding through the \suldiox~oxygens at the hot temperatures should have a visible effect on the radial distribution function (RDF) between \suldiox-oxygen and water-hydrogen. Figure \ref{fig:rdf} shows the RDF of the \suldiox~sulfur and oxygen atoms with their water bonding partners, water oxygen and hydrogen, respectively, at both cold (blue) and hot (green) temperatures. Both the first and second peaks of the O$_{SO_2}$-H$_{H_2O}$ RDF increase at the higher temperature, indicating a higher number of water hydrogens near the \suldiox~oxygens in both hydration shells. This corroborates the increased coordination through the \suldiox~oxygens to the surface waters at the higher temperature.

@@

The populations of the single-oxygen coordinations are similar at both temperatures, somewhat contradicting the results of the RDF. We expect an increase of all \suldiox-oxygen interactions with surface water hydrogens at the higher temperature, but the non-sulfur coordinations do not obey this trend. This indicates that the non-sulfur coordinations do not behave as most of the other coordinations, and their small contribution to the overall population in the distribution of Figure \ref{fig:bonding-coordinations} indicates that they are anomalous, or only briefly encountered during the simulations.

The same temperature effect of shifting population to higher bonding coordinations also manifests in the \suldiox~sulfur. The single-``S'' coordinations (``S'', ``SO'', ``SOO'', etc.) have similar populations at both temperatures, but the change from cold to hot temperature shifts the unbound \suldiox~coordinations up towards to the doubly-bound sulfur coordinations (i.e. ``SS'', ``SSO'', etc.). The increased temperature causes more binding to the \suldiox~by allowing for multiple bonds through both the \suldiox~sulfur, and through the oxygens. Thus, the hotter temperature allows for increased coordination of the \suldiox~to the water surface.

When the \suldiox~is on the cold water surface, the ``S'' and ``SO'' coordination populations are nearly equally represented. As the temperature increases to the hot system, the ``S'' coordination decreases, while the ``SO'' increases. We see from the RDFs that colder temperatures favor binding through the \suldiox~sulfur because of the larger first peak in the S$_{SO_2}$-O$_{H_2O}$ RDF. From this increase we expect the ``S''-coordinated \suldiox~to be more prevalent at cold temperatures, as is the case for the coordination distribution. However, the O$_{SO_2}$-H$_{H_2O}$ RDF indicates increased hydrogen bonding through the \suldiox~oxygens at higher temperatures manifested by a larger first peak in Figure \ref{fig:rdf} for the hot temperature. Thus, those \suldiox~bonding coordinations with at least a single bond to an oxygen and to the sulfur, (i.e. ``SO'', ``SSO'', ``SSOO'', etc.) are more prevalent at the high temperature. Overall, we find evidence from both the bonding coordination distribution, and from the RDFs to suggest a modified bonding behavior at different temperatures. At an increased temperature the surface \suldiox~binds more through its oxygens than the cold counterpart. However, like the \suldiox~in the cold temperature, it likely has a persistent interaction through the sulfur atom to a neighboring water oxygen throughout most of the simulated trajectories.
